Donor Responses to Civil Society Repression

### 2024-11-02 manuscript.qmd
Add Suparna's stuff

@@

## Foreign aid donors

A rich literature on the purposes of foreign aid explores a broad collection of reasons why states transfer money and resources to others. Aid agencies and international organizations face competing interests when deciding how to allocate resources, which projects to undertake, and which countries to work with. Donor agencies rhetorically declare normative justifications for engaging in development, with recipient-focused development goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011; @Kallman:2020]. In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy–oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017; @Schuller:2012], including being responsive to popular democratic pressure [@HeinrichKobayashiLong:2018], insulating themselves from regional instability [@Bermeo:2017], expanding export markets [@BarthelNeumayerNunnenkamp:2014; @Mitchell:2002], influencing recipient states' domestic policies [@Apodaca:2013], and projecting donor country values [@MorgensternBrown:2022].

### Donor responses and conditionality

@@

### Donor responses to civil society laws

@@

These shifts in funding patterns in response to restrictive civil society laws have had measurable effects on NGOs in recipient countries. Samantha Power, current administrator of USAID, has pledged to send 25% of USAID funding to local NGOs over the next few years [@Power:2021], though practical obstacles remain. New local partners take longer to vet and train compared to established INGO partners. Turning to local organizations entails greater risk, as domestic NGOs can be easily shut down based on the whims of the government. Additionally,  communities often distrust local groups that accept Western aid, especially those focused on democracy assistance [@Barkan:2012]. Donor preferences have also changed. Over the past two decades, bilateral donors have increasingly provided short-term grants for projects that have easily measurable, quantifiable outputs [@Bush:2015]. This model may work well for civic groups working on relatively tame causes or exclusively development-focused goals, but more contentious organizations facing government crackdown may need flexible funding that does not bring immediate results or does not have easily measurable goals [@Herrold:2020].

@@

>>>>

The experiences of organizations such as X are by no means a rare occurrence. While a large literature in law and society and political science focuses on the role of international law, domestic legal institutions, and legal mobilization that may reduce repression [@Klug:2005; @PowellStaton:2009; @Simmons:2009; @ConradRitter:2013], the last two decades have made it clear that states are also using law not to expand rights, but to further restrict rights and repress organizations [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @BakkeMitchellSmidt:2020; @GlasiusSchalkDeLange:2020; @BromleySchoferLonghofer:2020; @Chaudhry:2022]. Administrative crackdown, or the use of laws to create barriers to entry, funding, and advocacy by civil society groups has proliferated across the globe over the past few decades—both across democracies and autocracies [@Chaudhry:2022]. The latest annual report from CIVICUS Monitor, which assesses the state of civil society worldwide, paints a grim picture. As of 2023, a third of the world’s population now lives in countries with closed civic space, and 87% of the world’s population now lives in countries where the space available to civil society organizations (CSOs) is closed, repressed, or obstructed [@CIVICUSMonitor].

The growth in states using law to repress can be attributed to three main factors. First, administrative or legal crackdowns have many advantages over the use of violence. It allows governments to maintain a democratic facade abroad and limit the domestic backlash because citizens tend to view legal crackdown as regulation rather than repression [@Chaudhry:2022]. Anti-NGO laws rarely elicit international condemnation or threats to withdraw aid. States also learn about the efficacy of these legal strategies from their peers and subsequently implement them in their own country [@GlasiusSchalkDeLange:2020; @Gilbert:2020; @Chaudhry:2022]. Finally, these tactics are part of a broader trend of using legal means to subvert democratic institutions and norms [@LevitskyZiblatt:2018].

Administrative crackdown via anti-NGO laws that bar access to foreign funding and foreign technical assistance has been particularly insidious. Without access to foreign funds, many INGOs cannot continue their overseas operations. Even domestic NGOs may be heavily strapped for money. Most NGOs in the Global South require a constant stream of competitive grants from foreign donors to be able to implement specific projects [@Bush:2015]. Local groups tackling contentious issues may not be able to raise funds domestically, as publics may prefer donating to service organizations focusing on health, education, and poverty [@DupuyRonPrakash:2016; @Brechenmacher:2017]. Even local philanthropists may be deterred from giving to organizations focusing on contentious causes such as advocacy, media freedom, and anti-corruption initiatives due to poor tax incentives or a fear of retribution.

The effects of such repression are visible in many countries. For instance, the 2009 Ethiopian Charities and Societies Proclamation Act stipulated that NGOs working on any rights issues in the country must acquire 90% of their funding from domestic sources. It also prohibits organizations from spending more than 30% of their budget on "administrative costs," which the act does not define, but could be interpreted to include the provision of free legal aid, advocacy, and other activities essential to the missions of rights groups [@Brechenmacher:2017, 69]. This law dealt a devastating blow to Ethiopian civil society, as most domestic NGOs working on human rights relied almost exclusively on foreign aid [@Brechenmacher:2017]. Within just two years, the total number of local NGOs reduced by 25%, and by 90% among human rights NGOs [@DupuyRonPrakash:2016]. INGOs such as Mercy Corps and Action for Development abandoned their conflict resolution work, while Human Rights Watch, Amnesty International, and the International Federation for Human Rights were prevented from opening field offices in the country. Only 10% of the 125 human rights groups in operation at the time of the law’s passage registered under the new requirement [@DupuyRonPrakash:2016].

@@

# Foreign aid donors

A rich literature on the purposes of foreign aid explores a broad collection of reasons why states transfer money and resources to others. Aid agencies and international organizations face competing interests when deciding how to allocate resources, which projects to undertake, and which countries to work with. Donor agencies rhetorically declare normative justifications for engaging in development, with recipient-focused development goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011; @Kallman:2020]. In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy-oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017; @Schuller:2012], including being responsive to popular democratic pressure [@HeinrichKobayashiLong:2018], insulating themselves from regional instability [@Bermeo:2017], expanding export markets [@BarthelNeumayerNunnenkamp:2014], influencing recipient states' domestic policies [@Apodaca:2013], and projecting donor country values [@MorgensternBrown:2022].

## Donor responses and conditionality

@@

## State and foundation responses

@@

These shifts in funding patterns in response to restrictive civil society laws have had measurable effects on NGOs in recipient countries. Samantha Power, current administrator of USAID, has pledged to send 25% of USAID funding to local NGOs over the next few years [@Power:2021], though practical obstacles remain. New local partners take longer to vet and train compared to established INGO partners. Turning to local organizations entails greater risk, as domestic NGOs can be easily shut down based on the whims of the government. Additionally, communities often distrust local groups that accept Western aid, especially those focused on democracy assistance [@Barkan:2012]. Donor preferences have also changed. Over the past two decades, bilateral donors and foundations have increasingly provided short-term grants for projects that have easily measurable, quantifiable outputs [@Bush:2015]. This model may work well for civic groups working on relatively tame causes or exclusively development-focused goals, but more contentious organizations facing government crackdown may need flexible funding that does not bring immediate results or does not have easily measurable goals [@Herrold:2020].

@@

# Local solutions

Due to the reduced access to (or prohibition on) foreign funds, many local NGOs are also exploring alternative funding models. Some NGOs have adapted by engaging in diversification: they create additional affiliate organizations to ensure their survival [@ToeplerPapeBenevolenski:2020]. They may switch some of their work to another form of organization—including for-profits, which are not subject to the same legal restrictions. Creating commercial organizations affiliated with an NGO also allows financial assets to be transferred to an organization that is not subject to the same anti-NGO restrictions. In Russia, many formalized environmental NGOs have employed informal models of organization and deregistered with the state due to political and financial constraints. They have also diversified and are now working with businesses to accomplish their goals. For instance, the Ecological Union in St. Petersburg offers private green certification to businesses that undergo a third-party audit of their manufacturing and retail processes [@SundstromHenrySperling:2022].

A growing number of organizations in the Global South are also investing in community philanthropy, women's funds, and developing social enterprises that rely on business revenue. Community foundations facilitate democratic and transparent decision making by the local community, which decides who gets grants, and how much. This approach allows local groups to respond to (and be held accountable to) local priorities, which also helps overcome many criticisms of international donor aid [@Murad:2014].

Developing social enterprises is another alternative funding model for NGOs. Such enterprises are privately owned, for-profit, nonprofit, or hybrid organizations with social objectives. They often advance their cause using a business model and then reinvest profits in the community or the nonprofit. For instance, Bangladesh has some of the most notable social enterprise nonprofit organizations, including the Bangladesh Rural Advancement Committee (BRAC) and the Grameen Bank. BRAC's programs address a wide variety of issues including health, education, craft markets, milk and cattle farming, and micro-credit. During the organization's initial stage of development, it reinvested half of its commercial pro ts back into the enterprise and the other half into its nonprofit programming [@ChoSultana:2015]. Such alternative funding models can thus generate sustainable income while achieving non-profit goals of socioeconomic and community development.

# Continued challenges from anti-NGO rhetoric

Foreign donors already face many criticisms from grantees, including implementing models without taking local contexts into account and prioritizing projects that are more easily quantifiable and regime compatible. In an environment of closing civic space, despite many of the adaptations discussed above, NGOs—especially those receiving grants from foreign donors—face even bigger risks. In a 2011 survey of 1,473 NGO respondents that received funds from 14 foreign donor agencies, 37% felt the greatest risk of receiving democracy assistance was being labeled an "agent" or "stooge" of outside forces; 48% of respondents lived in countries categorized as "Not Free" by Freedom House [@Barkan:2012]. In a 2016 survey of over 1,000 activists from Colombia, Egypt, India, Kenya, Russia, Sudan, Turkey, Uganda, Ukraine, and Venezuela, many said that foreign actors often contributed to their security or safety concerns by focusing on digital security to the exclusion of more pressing physical security needs and by failing to take measures to protect partners, including when they have been targeted or arrested [@Miller-Dawkins:2017].

Such risk does not necessarily just imply threats to physical or legal safety. Many governments around the world also utilize negative rhetoric that aims to sow distrust between these groups and the communities they serve. In Hungary, government officials declared in regards to NGOs that they wanted to "sweep them out of the country," while asking the intelligence services to investigate all NGOs receiving money from the Hungarian-born financier George Soros [@Reuters:2017]. In rhetoric similar to that in Hungary, where government billboards warn "Don't let George Soros have the last laugh," the Trump administration accused protestors opposing Brett Kavanaugh’s confirmation to the Supreme Court of being funded by wealthy liberal donors like Soros [@Choi:2018].

Recent research shows that NGOs do consider government rhetoric when making operational decisions. In a survey experiment of NGOs across Cambodia, Uganda, and Serbia, researchers found that negative NGO rhetoric significantly reduces NGO leaders’ willingness to work in a community. It also rescues NGOs’ preferences for partnering with local government authorities [@HollerbauerNagawaRobertson:2024]. Thus, NGOs prefer to avoid working in communities not just where government authorities repress these groups, but also in areas where authorities use anti-NGO rhetoric. While donor responses have largely focused on navigating and adapting to repression, the use of such rhetoric raises broader questions about research and policy on the nonprofit sector: how is such rhetoric changing public attitudes and engagement with these groups? Given that the looming threat of repression, and even anti-NGO rhetoric can exacerbate burnout rates [@JoscelyneKnuckeySatterthwaite:2015], what are organizations doing to support human rights advocates, and what kinds of interventions can help support human rights activists and professionals?

In response to these concerns, organizations such as Defend Defenders, Frontline Defenders, Peace Brigades International, Civil Rights Defenders, and Protection International provide physical security, legal and digital support, and mental health resources to activists at risk. Regional initiatives such as the Mesoamerican Women Human Rights Defenders Initiative, Forum Asia, and the Euro-Mediterranean Foundation of Support for Human Rights Defenders also help provide formal protection mechanisms for activists.

In many cases, foreign donors have also allied with local NGOs, and provided training to local groups to help defeat anti-NGO bills. In Kyrgyzstan, when legislators introduced a draft bill outlining NGO restrictions, Freedom House collaborated with local NGOs to lobby the government. They warned legislators that the proposed Russian-style "Foreign Agents" law would adversely affect charitable and humanitarian NGOs and that these restrictions would "lead to a reduction in the number of social services that the population of the country desperately need" [@CheesemanDodsworth:2023, 624]. At the international level, the US, EU, and DAC-EU members—which provide a quarter to a third of official development assistance to Kyrgyzstan—also made their concerns known to legislators [@CheesemanDodsworth:2023]. The bill was eventually defeated in 2016. Kyrgyz legislators who voted against the bill cited the need for foreign funding: "We get financial assistance from [international organizations] in many fields including healthcare, education, and agriculture among others. We need this money" [@Lelik:2016].

Local NGOs similarly partnered with INGOs in Kenya; their campaign emphasized that 8,500 organizations in the country rely heavily on foreign donors and that the proposed 15% cap on foreign funding would have devastating consequences for the NGO sector. During the subsequent reading of amendments to the NGO bill, several legislators explained their rejection of the bill in terms of the risks for the socioeconomic development of Kenya and their constituents [@Berger-KernHetzWagner:2021]. These examples highlight how donors can use their leverage, especially their role in social service provision, to successfully pushback against repressive NGO laws.

# Conclusion

Donors, especially international ones, have long faced criticisms by practitioners, policymakers and academics alike, arguing that they replicate global hierarchies and are unresponsive to the needs of local communities. But increasing civic space crackdown, especially in many established democracies in the West, means that when such countries, or donors in these countries speak out or take actions against civil society restrictions abroad, they may have less credibility. For instance, the U.S. government has teetered dangerously close to the rhetoric and practices shown by many other governments repressing civil society. In June 2018, Nikki Haley, U.S. Ambassador to the United Nations (UN), blamed U.S. withdrawal from the UN Human Rights Council on NGOs, a move that Sarah Morgan, Washington Director of Human Rights Watch, called "astounding." This is intrinsic to the reasons why so far, international actors have struggled to mount a coherent campaign.

While examples of Kenya and Kyrgyzstan discussed above show that donors and local activists can successfully push back against anti-NGO laws, a more systematic approach is needed. To tackle this phenomenon more systematically, Western states need a more effective coordinated response that engages not just aid agencies, donors, and international NGOs, but also engages businesses, development communities, diplomatic actors and local populations to deal with larger structural issues that exacerbate government crackdown on civil society. Without such an effort, repression of civil society organizations may be the canary in the coal mine that sets the groundwork for future democratic erosion.
